---
title: "Introduction to ggplot2"
description: |
  A discussion of ggplot2 terminology, and an example of iteratively refining a
  simple scatterplot.
author: Kris Sankaran, adapted by Cameron Jones
date: Summer 2025
  
output: rmdformats::robobook
---

```{r, echo = FALSE}
library("knitr")
knitr::opts_chunk$set(cache = FALSE, message = FALSE, warning = FALSE, echo = TRUE)
```

# Introduction

[ggplot2](https://ggplot2.tidyverse.org/) is an R implementation of the _Grammar
of Graphics_. The idea is to define the basic "words" from which visualizations
are built, and then let users compose them in original ways. This is in contrast
to systems with prespecified chart types, where the user is forced to pick from
a limited dropdown menu of plots. Just like in ordinary language, the creative
combination of simple building blocks can support a very wide range of
expression.

These are libraries we'll use in this lecture.

```{r}
library(tidyverse)
library(dslabs)
library(ggrepel)
library(scales)
```

# Iterative Visualization

We're going to create this plot in these notes.

```{r, echo = FALSE}
data(murders)
r <- murders %>% 
  summarize(rate = sum(total) /  sum(population)) %>%
  pull(rate)

ggplot(murders, aes(x = population, y = total)) +
  geom_abline(intercept = log10(r), linewidth = 0.4, col = "#b3b3b3") +
  geom_text_repel(aes(label = abb), segment.size = 0.2) + # I moved it up so that the geom_points appear on top of the lines
  geom_point(aes(col = region)) +
  scale_x_log10(labels = unit_format(unit = "million", scale = 1e-6)) + # used to convert scientific notation to readable labels
  scale_y_log10() +
  scale_color_brewer(palette = "Set2") +
  labs(
    x = "Population (log scale)",
    y = "Total number of murders (log scale)",
    color = "Region",
    title = "US Gun Murders in 2010"
  ) +
  theme_bw() +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank()
  )
```

Every ggplot2 plot is made from three components,

* Data: This is the `data.frame` that we want to visualize.
* Geometry: These are the types of visual marks that appear on the plot.
* Aesthetic Mapping: This links the data with the visual marks.

# The Data

Let’s load up the data. Each row is an observation, and each column is an
attribute that describes something about that observation. This is important because each mark that you see on a ggplot -- a line, a point, a tile, ... -- had to start out as a row within an R `data.frame` (with the exception of some one-variable plots like boxplots) The visual properties of the mark (e.g., color)
are determined by the values along columns.

Here's an example of the data above in tidy format,

```{r}
data(murders)
head(murders)
```

This is one example of how the same information might be stored in a non-tidy
way, making visualization much harder.

```{r}
non_tidy <- data.frame(t(murders))
colnames(non_tidy) <- non_tidy[1, ]
non_tidy <- non_tidy[-1, ]
non_tidy[, 1:6]
```

Often, one of the hardest parts in making a ggplot2 plot is not coming up with
the right ggplot2 commands, but reshaping the data so that it’s in a tidy
format.

# Geometries

The words in the grammar of graphics are the geometry layers. We can associate
each row of a data frame with points, lines, tiles, etc., just by referring to
the appropriate geom in ggplot2. A typical plot will compose a chain of layers
on top of a dataset,

> ggplot(data) + [layer 1] + [layer 2] + ...

For example, by deconstructing the plot above, we would expect to have point and
text layers. For now, let's just tell the plot to put all the `geom`'s at the
origin.

```{r}
ggplot(murders) +
  geom_point(x = 0, y = 0) +
  geom_text(x = 0, y = 0, label = "test")
```

You can see all the types of `geoms` in the [cheat
sheet](https://github.com/rstudio/cheatsheets/blob/master/data-visualization-2.1.pdf).

# Aesthetic Mappings

Aesthetic mappings make the connection between the data and the geometry. It's
the piece that translates abstract data fields into visual properties. Analyzing
the original graph, we recognize these specific mappings,

* State Population; $x$-axis coordinate
* Number of murders; $y$-axis coordinate
* Geographical region; color

To establish these mappings, we need to use the `aes` function. Notice that
column names don't have to be quoted, or preceded with `murders$` -- ggplot2 knows to refer back to the data frame we specified in the original `ggplot` call, which was `murders`.

```{r}
ggplot(murders) +
  geom_point(aes(x = population, y = total, col = region))
```

The original plot used a log-scale. To transform the x and y axes, we can use
[scales](https://ggplot2.tidyverse.org/reference/index.html#section-scales).

```{r}
ggplot(murders) +
  geom_point(aes(x = population, y = total, col = region)) +
  scale_x_log10() +
  scale_y_log10()
```

One nuance is that scales aren't limited to $x$ and $y$ transformations. They
can be applied to modify any relationship between a data field and its
appearance on the page. For example, this changes the mapping between the region
field and circle color.

```{r}
ggplot(murders) +
  geom_point(aes(x = population, y = total, col = region)) +
  scale_x_log10() +
  scale_y_log10() +
  scale_color_manual(values = c("#6a4078", "#aa1518", "#9ecaf8", "#50838c"))
```

A problem with this graph is that it doesn't tell us which state each point
corresponds to. For that, we'll need text labels. We can encode the coordinates
for these marks again using `aes`, but this time within a `geom_text` layer.

```{r}
ggplot(murders) +
  geom_point(aes(x = population, y = total, col = region)) +
  geom_text(
    aes(x = population, y = total, label = abb),
    nudge_x = 0.08 # what would happen if I remove this?
  ) +
  scale_x_log10() +
  scale_y_log10()
```

Note that each type of layer uses different visual properties to encode the data
-- the argument `label` is only available for the `geom_text` layer. You can see
which aesthetic mappings are required for each type of `geom` by checking that
`geom`'s documentation page, under the Aesthetics heading.

It's usually a good thing to make your code as concise as possible. For ggplot2,
we can achieve this by sharing elements across `aes` calls (e.g., not having to
type `population` and `total` twice). This can be done by defining a "global"
aesthetic, putting it inside the initial `ggplot` call.

```{r}
ggplot(murders, aes(x = population, y = total)) +
  geom_point(aes(col = region)) +
  geom_text(aes(label = abb), nudge_x = 0.08) +
  scale_x_log10() +
  scale_y_log10()
```

# Finishing Touches

How can we improve the readability of this plot? You might already have ideas,

1. Prevent labels from overlapping. It's impossible to read some of the state names.
2. Add a line showing the national rate. This serves as a point of reference,
allowing us to see whether an individual state is above or below the national
murder rate.
3. Give meaningful axis / legend labels and a title.
4. Move the legend to the top of the figure. Right now, we're wasting a lot of
visual real estate in the right hand side, just to let people know what each
color means.
5. Use a better color theme.

For 1., the `ggrepel` package offers a function `geom_text_repel` which finds better positions for the labels, drawing links
when necessary. (The downside to )

```{r}
ggplot(murders, aes(x = population, y = total)) +
  geom_text_repel(aes(label = abb), segment.size = 0.2) + # Moved this layer here so the points are on top of the connecting segments
  geom_point(aes(col = region)) +
  scale_x_log10() +
  scale_y_log10()
```

For 2., let's first compute the national murder rate,

```{r}
r <- murders %>% 
  summarize(rate = sum(total) /  sum(population)) %>%
  pull(rate)
r
```

Now, we can use this as the slope in a `geom_abline` layer, which encodes a slope and intercept as a line on a graph.

```{r}
ggplot(murders, aes(x = population, y = total)) +
  geom_abline(intercept = log10(r), linewidth = 0.4, col = "#b3b3b3") +
  geom_text_repel(aes(label = abb), segment.size = 0.2) +
  geom_point(aes(col = region)) +
  scale_x_log10() +
  scale_y_log10()
```

For 3., we can add a `labs` layer to write labels and a `theme` to reposition the legend. I used `unit_format` from the `scales` package to change the scientific notation in the $x$-axis labels to something more readable.

```{r}
ggplot(murders, aes(x = population, y = total)) +
  geom_abline(intercept = log10(r), linewidth = 0.4, col = "#b3b3b3") +
  geom_text_repel(aes(label = abb), segment.size = 0.2) +
  geom_point(aes(col = region)) +
  scale_x_log10(labels = unit_format(unit = "million", scale = 1e-6)) + # used to convert scientific notation to readable labels
  scale_y_log10() +
  labs(
    x = "Population (log scale)",
    y = "Total number of murders (log scale)",
    color = "region",
    title = "US Gun Murders in 2010"
  ) +
  theme(legend.position = "top")
```

For 5., I find the gray background with reference lines a bit distracting. We
can simplify the appearance using `theme_bw`. I also like the colorbrewer
palette, which can be used by calling a different color scale.

```{r}
ggplot(murders, aes(x = population, y = total)) +
  geom_abline(intercept = log10(r), linewidth = 0.4, col = "#b3b3b3") +
  geom_text_repel(aes(label = abb), segment.size = 0.2) +
  geom_point(aes(col = region)) +
  scale_x_log10(labels = unit_format(unit = "million", scale = 1e-6)) +
  scale_y_log10() +
  scale_color_brewer(palette = "Set2") +
  labs(
    x = "Population (log scale)",
    y = "Total number of murders (log scale)",
    color = "Region",
    title = "US Gun Murders in 2010"
  ) +
  theme_bw() +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank()
  )
```

# Beyond Finishing Touches

Almost anything can be customized in a ggplot; there is far too much for us to A) list every possible way things can be customized or B) expect you to know them all, even if we could list them.

While you visualize, we highly encourage you to explore ggplot customization on your own. Learning how to Google and StackOverflow tiny ggplot tweaks is a huge part of visualization that won't show up in any course learning outcome. Thoughts like "how do I get rid of the gray background around the points in the legend?" or "how do I move the X axis title a little further away from the plot?" or "how do I get rid of the minor gridlines between the major ones?" are perfect candidates for things you can learn with a quick search.

In that spirit, here are some not-straightforward bonus exercises for the curious. These encourage you to look at your graphics more carefully and start your mental collection of customizations!

* Re-order the order of regions in the legend. Hint: Reset the factor levels in
the `region` field of the `murders` data.frame.
* Only show labels for a subset of states that are far from the national rate.
Hint: Filter the `murders` data.frame, and use a `data` field specific to the
`geom_text_repel` layer.
* Move "Population (log scale)" a little further down, away from the plot. 
* Increase the size of the circles in the legend, but not in the plot. Hint: Use `override.aes`
within a `guide`.

See the R Markdown file on the course GitHub for solutions!

```{r, echo = FALSE}

extreme_states_for_labeling = murders %>% 
  mutate(rate = total/population, distance = abs(rate - r)) %>%
  slice_max(distance, n = 15)

ggplot(murders, aes(x = population, y = total)) +
  geom_abline(intercept = log10(r), linewidth = 0.4, col = "#b3b3b3") +
  geom_text_repel(data = extreme_states_for_labeling, aes(label = abb), segment.size = 0.2) +
  geom_point(aes(col = factor(region, levels = c("West", "North Central", "South", "Northeast")))) + # Reorder factor levels
  scale_x_log10(labels = unit_format(unit = "million", scale = 1e-6)) +
  scale_y_log10() +
  scale_color_brewer(palette = "Set2") +
  guides(colour = guide_legend(override.aes = list(size=6))) + # Increase size of circles in legend
  labs(
    x = "Population (log scale)",
    y = "Total number of murders (log scale)",
    color = "Region",
    title = "US Gun Murders in 2010"
  ) +
  theme_bw() +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank(),
    axis.title.x = element_text(margin = margin(t = 8)) # Move x axis title down; "give 8 units of room on top"
  )
```
