---
title: "Compound Figures"
description: |
  Showing different variables across subpanels.
author: Kris Sankaran, adapted by Cameron Jones
date: Summer 2025
  
output: rmdformats::robobook
---

```{r, echo = FALSE, warning = FALSE, message = FALSE}
library(tidyverse)
library(knitr)
knitr::opts_chunk$set(cache = TRUE, message = FALSE, warning = FALSE, echo = FALSE, fig.height = 5, retina = 4)
th <- theme_minimal() + 
  theme(
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "#f7f7f7"),
    panel.border = element_rect(fill = NA, color = "#0c0c0c", size = 0.6),
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16),
    legend.position = "bottom"
  )
theme_set(th)
```

# Why Compound Plots?

Faceting is useful whenever we want different **rows** of the data to appear in
different panels. What if we want to compare different columns, or work with
several datasets? A more general alternative is to use compound plots. The idea
is to construct plots separately and then combine them only at the very end.

The main advantage of compound plots is that individual panels can be
tailored to specific visual comparisons, but relationships across panels can
also be studied. For example, the plot below shows change in the total number
and composition of undergraduate majors over the last few decades. In principle,
the same information could be communicated using a stacked area plot
(`geom_area`). However, comparing the percentages for 1970 and 2015 is much more
straightforward using a line plot, and we can still see changes in the overall
number of degrees using the area plot.
    
```{r, fig.width = 10, fig.height = 4}
    library(tidyverse)
    library(patchwork)
    library(scales)
    
    degrees <- read_csv("https://raw.githubusercontent.com/krisrs1128/stat436_s23/main/data/degrees.csv") %>%
        filter(field %in% c("Business", "Health professions and related programs", "Social sciences and history", "Psychology", "Education"))
    
    p <- list()
    p[["trend"]] <- degrees %>%
      group_by(year) %>%
      summarise(total = sum(count)) %>%
      ggplot() +
        geom_area(aes(year, total)) +
        scale_x_continuous(expand = c(0, 0)) +
        scale_y_continuous(expand = c(0, 0, 0.1, 0), labels = label_number(scale_cut = cut_short_scale())) +
        labs(x = NULL, y = "degrees awarded")
    
    delta_data <- degrees %>%
      filter(year %in% c(1971, 2015))
    
    p[["delta"]] <- ggplot(delta_data, aes(year_str, perc)) +
      scale_x_discrete(expand = c(0.05, 0.01, 0, 0.9)) +
      scale_y_continuous(labels = label_percent()) +
      geom_point() +
      geom_text(
        data = delta_data %>% filter(year == 2015), 
        aes(label = str_wrap(field, 20)),
        size = 3.5, nudge_x = 0.04, hjust = 0
      ) +
      geom_line(aes(group = field)) +
      labs(x = NULL, y = "proportion of degrees")
    
    p[["trend"]] + p[["delta"]] +
      plot_layout(widths = c(0.65, 0.35))
```
    
For reference, here is a non-compound display of the same information.
        
```{r}
    ggplot(degrees) +
      geom_area(aes(year, count, fill = field)) +
        scale_x_continuous(expand = c(0, 0)) +
        scale_y_continuous(expand = c(0, 0, 0.1, 0), labels = label_number(scale_cut = cut_short_scale())) +
        scale_fill_brewer(palette = "Set2")
```
    
# Considerations  
    
There are a few considerations that can substantially improve the quality of
a compound plot,

* Consistent visual encodings for shared variables
* Clear, but unobtrusive annotation
* Proper alignment in figure baselines
    
We will discuss each point separately.
 
## Consistent Encodings 
    
The figures below are compound plots of a dataset of athlete physiology. They
are very similar, but the second is better because it enforces a more strict
consistency in encodings across panels. Specifically, the male / female variable
is (1) encoded using the same color scheme across all panels and (2) ordered so
that female repeatedly appears on the right of male.
    
```{r}
    athletes <- read_csv("https://raw.githubusercontent.com/krisrs1128/stat436_s23/main/data/athletes.csv") %>%
      filter(sport %in% c("basketball", "field", "rowing", "swimming", "tennis", "track (400m)"))
    p <- list()
    
    bar_plot <- athletes %>%
      count(sex) %>%
      mutate(sex = factor(sex, levels = c("m", "f"))) %>%
      ggplot() +
      geom_bar(aes(sex, n), stat = "identity") +
      scale_y_continuous(expand = c(0, 0))
    scatter_plot <- ggplot(athletes) +
      geom_point(aes(rcc, wcc, col = sex)) +
      scale_color_brewer(palette = "Set1")
    box_plot <- ggplot(athletes) +
      geom_boxplot(aes(sport, pcBfat, fill = sex))
    
    (bar_plot + scatter_plot) / box_plot
```
    
The improved, visually consistent approach is given below.
    
```{r}
    athletes <- athletes %>%
      mutate(sex = recode(sex, "m" = "male", "f" = "female"))
    
    bar_plot <- athletes %>%
      count(sex) %>%
      ggplot() +
      geom_bar(aes(sex, n, fill = sex), stat = "identity") +
      scale_y_continuous(expand = c(0, 0)) +
      scale_fill_brewer(palette = "Set1") +
      labs(y = "number")
    scatter_plot <- ggplot(athletes) +
      geom_point(aes(rcc, wcc, col = sex)) +
      scale_color_brewer(palette = "Set1") +
      theme(legend.position = "none") +
      labs(x = "RBC count", y = "WBC Count")
    box_plot <- ggplot(athletes) +
      geom_boxplot(aes(sport, pcBfat, col = sex, fill = sex), alpha = 0.5) +
      scale_color_brewer(palette = "Set1") +
      scale_fill_brewer(palette = "Set1") +
      theme(legend.position = "none") +
      labs(y = "% body fat", x = NULL)
    
    (bar_plot + scatter_plot) / box_plot +
      plot_layout(guides = "collect") &
      plot_annotation(theme = theme(legend.position = "bottom"))
```

## Annotation
  
Effective annotation can be used to refer to different subpanels of the data
without drawing too much attention to itself. Labels should be visible but
subtle -- not too large, similar fonts as the figures, and logically ordered
(e.g. (a) on top left). A nice heuristic is to think of these annotations like page
numbers. They are useful for making references, but aren't something that is
actively read.
    
```{r}
    bar_plot <- bar_plot + ggtitle("a")
    scatter_plot <- scatter_plot + ggtitle("b")
    box_plot <- box_plot + ggtitle("c")
    
    (bar_plot + scatter_plot) / box_plot +
      plot_layout(guides = "collect") &
      plot_annotation(theme = theme(legend.position = "bottom"))
```
  
For alignment, we will want figure baselines / borders to be consistent.
  Misalignment can be distracting. This is primarily a problem when compound plots
  are made manually; for example, by pasting screenshots together into a document.
  
Using `patchwork` below, we won't have this issue.
      
# Patchwork

The `patchwork` package does not exist to help you *make* the individual figures, rather to effectively *combine* existing figures.

This package creates a simple syntax for combining figures `p1` and `p2`,

    * `p1 + p2` concatenates two figures horizontally
    * `p1 / p2` concatenates two figures vertically
    
This idea is simple, but becomes very powerful once we realize that we can define a whole algebra on plot layouts,
    
    * `p1 + p2 + p3` concatenates three figures horizontally
    * `p1 / p2 / p3` concatenates three figures vertically
    * `(p1 + p2) / p3` Concatenates the first two figures horizontally, and places the third below both.
    * ...

    
```{r}
  athletes <- read_csv("https://raw.githubusercontent.com/krisrs1128/stat436_s23/main/data/athletes.csv") %>%
      filter(sport %in% c("basketball", "field", "rowing", "swimming", "tennis", "track (400m)")) %>%
      mutate(sex = recode(sex, "m" = "male", "f" = "female"))
    

    bar_plot <- ggplot(count(athletes, sex)) +
      geom_bar(aes(sex, n, fill = sex), stat = "identity") +
      scale_y_continuous(expand = c(0, 0)) +
      scale_fill_brewer(palette = "Set1") +
      labs(y = "number")
    
    scatter_plot <- ggplot(athletes) +
      geom_point(aes(rcc, wcc, col = sex)) +
      scale_color_brewer(palette = "Set1") +
      theme(legend.position = "none") +
      labs(x = "RBC count", y = "WBC Count")
    
    box_plot <- ggplot(athletes) +
      geom_boxplot(aes(sport, pcBfat, col = sex, fill = sex), alpha = 0.5) +
      scale_color_brewer(palette = "Set1") +
      scale_fill_brewer(palette = "Set1") +
      theme(legend.position = "none") +
      labs(y = "% body fat", x = NULL)
```
  
 Now, we use `patchwork` to combine the subplots using the different
combinations discussed above.
    
```{r, echo = TRUE}
    bar_plot + scatter_plot + box_plot
    bar_plot / scatter_plot / box_plot
    (bar_plot + scatter_plot) / box_plot
```

A corollary of using the same encodings across panels is that it should be
possible to share legends across the entire compound figure. This is most
concisely done by setting `plot_layout(legend = "collect")`. For example,
compare the athlete physiology dataset without and with the collected legends:

Here's the version where every panel has its own legend for the same aesthetic:
    
```{r}
    (bar_plot + scatter_plot + theme(legend.position = "left")) / box_plot + theme(legend.position = "left") # turns legends back on
```

And here's the version where the legends are collected into one:

```{r}
    (bar_plot + scatter_plot) / box_plot +
          plot_layout(guides = "collect") &
          plot_annotation(theme = theme(legend.position = "bottom"))
```
  
 For annotation, we can add a title to each figure individually using
`ggtitle()`, before they are combined into the compound figure. The size and
font of the titles can be adjusted by using the `theme(title =
element_text(...))` option. For example, the code below adds the a - c titles
for each subpanel.
    
```{r}
    bar_plot <- bar_plot + ggtitle("a")
    scatter_plot <- scatter_plot + ggtitle("b")
    box_plot <- box_plot + ggtitle("c")
    
    (bar_plot + scatter_plot) / box_plot +
      plot_layout(guides = "collect") &
      plot_annotation(theme = theme(legend.position = "bottom", title = element_text(size = 10)))
```
    
 Patchwork handles alignment in the background, but sometimes we might want to
have control over the relative sizes of different panels. For this, we can again
use the `plot_layout` function, this time using the height and width arguments.
For example, these two examples change the relative widths and heights of the layout components.

```{r, echo = TRUE}
    (bar_plot + scatter_plot + plot_layout(widths = c(1, 3))) / box_plot +
      plot_layout(guides = "collect")

    (bar_plot + scatter_plot) / box_plot +
      plot_layout(guides = "collect", heights = c(1, 3))
```
