---
title: "Untitled"
output: html_document
date: "`r Sys.Date()`"
---

```{r, echo = FALSE}
knitr::opts_chunk$set(warnings = FALSE, message = FALSE)
```

```{r}
library(sf)
library(tmap)
library(tidyverse)
library(ceramic)

# code to set a clean theme
th <- theme_minimal() + 
  theme(
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "#F7F7F7"),
    panel.border = element_rect(fill = NA, color = "#0c0c0c", size = 0.6),
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16),
    legend.position = "bottom"
  )
theme_set(th)
```

[Glacial Lakes]. The data at this [link](https://raw.githubusercontent.com/krisrs1128/stat436_s23/main/data/GL_3basins_2015.topojson) contain labels of glacial lakes in the Hindu Kush Himalaya, created during an ecological survey in 2015 by the International Centre for Integrated Mountain Development.

```{r}
lakes <- read_sf("https://raw.githubusercontent.com/krisrs1128/stat436_s23/main/data/GL_3basins_2015.topojson") |>
  mutate(log_area = log(Area))

```

a. How many lakes are in this dataset? What are the latitude / longitude
coordinates of the largest lakes in each sub-basin?

```{r}
top_lakes = lakes %>% 
  group_by(Sub_Basin) %>% 
  slice_max(Area, n = 1)
```

b. Plot the polygons associated with each of the lakes identified in step
(a). 

*Hint: You may find it useful to split lakes across panels using the `tm_facets` function. If you use this approach, make sure to include a scale with `tm_scale_bar()`, so that it is clear that lakes have different sizes.*

```{r}
tm_shape(top_lakes) +
  tm_polygons(col = "Area") +
  tm_facets("Sub_Basin") +
  tm_scale_bar()
```

```{r}
ggplot(top_lakes) +
  geom_sf(aes(fill = Area))
```

Advantages of `tm` over `ggplot`:
- Much easier to do complex geographic operations, like faceting or `tm_scale_bar()`
- Works with vector or raster (`geom_sf` only works with vector, but that's more common)

Advantage of `geom_sf()`:
- Interfaces with the rest of ggplot2 seamlessly

c. Visualize all lakes with latitude between 28.2 and 28.4 and with longitude between 85.8 and 86. 

```{r}
# Reminder to self: bbox = c(long_min, lat_min, long_max, lat_max)
# bbox = c(85.8, 28.2, 86, 28.4)

tm_shape(lakes, bbox = c(85.8, 28.2, 86, 28.4)) +
  tm_polygons(col = "Area")

ggplot(lakes) +
  geom_sf(aes(fill = Area)) +
  xlim(c(85.8, 86)) +
  ylim(c(28.2, 28.4))
```

