---
title: "Untitled"
output: html_document
date: "`r Sys.Date()`"
---

```{r, echo = FALSE}
knitr::opts_chunk$set(warnings = FALSE, message = FALSE)
```


[Interactive German Traffic] This problem will revisit the previous problem
from an interactive point of view. We will build a visualization that helps
users explore daily traffic patterns across multiple German cities, using
interactivity to help users navigate the collection. We will need additional
features related to the day of the week for each timepoint, created by the
`wday` function below,

```{r}
library(tidyverse)
library(lubridate)
library(shiny)
theme_set(theme_bw())

traffic <- read_csv("https://uwmadison.box.com/shared/static/x0mp3rhhic78vufsxtgrwencchmghbdf.csv") |>
 mutate(day_of_week = wday(date))
```

a. Design and implement a Shiny app that allows users to visualize traffic
over time across selected subsets of cities. Make sure that it is possible
to view data from more than one city at a time. It is not necessary to label
the cities within the associated figure.

We first define a function that, when given a subset of cities, draws a line
plot.

```{r, eval = FALSE}
plot_traffic <- function(df) {
  ggplot(df) +
    geom_line(aes(date, value, group = name)) +
    labs(x = "Date", y = "Traffic") +
    theme(axis.title = element_text(size = 20))
}
```

Our design will update a time series plot of all the cities every time a
dropdown menu is updated. We will allow multiple cities to be selected
simultaneously. Specifically, our UI has an input for choosing cities and
displays the line plot as an output. Our server recognizes changes in the
choice of cities, filters the data to that subset, and then draws the updated
time series.
    
```{r, eval = FALSE}
ui <- fluidPage(
  selectInput("city", "City", unique(traffic$name), multiple = TRUE),
  plotOutput("time_series")
)

server <- function(input, output) {
  output$time_series <- renderPlot({
    traffic |>
      filter(name %in% input$city) |>
      plot_traffic()
  })
}

shinyApp(ui, server)
```


b. Introduce new inputs to allow users to select a contiguous range of days
of the week. For example, the user should have a way of zooming into the
samples taken within the Monday - Wednesday range.

We use nearly the same design except that a new slider input is provided for
choosing days of the week. When a range of days is chosen, then the time
series will show only that range for the currently selected cities.

```{r, eval = FALSE}
ui <- fluidPage(
  selectInput("city", "City", unique(traffic$name), multiple = TRUE),
  sliderInput("day_of_week", "Days", 2, 7, c(2, 7)),
  plotOutput("time_series")
)

server <- function(input, output) {
  output$time_series <- renderPlot({
    traffic |>
      filter(
        name %in% input$city, 
        day_of_week >= input$day_of_week[1] & day_of_week <= input$day_of_week[2]
      ) |>
      plot_traffic()
  })
}

shinyApp(ui, server)
```
  
```{r}

library(shiny)
library(tidyverse)
library(lubridate)
library(plotly)

movies <- read_csv("https://raw.githubusercontent.com/krisrs1128/stat479_s22/main/_posts/2022-02-10-week04-03/apps/data/movies.csv") %>%
  mutate(
    date = as_date(Release_Date, format = "%b %d %Y"),
    year = year(date),
    Major_Genre = fct_explicit_na(Major_Genre),
    MPAA_Rating = fct_explicit_na(MPAA_Rating),
  )

genres <- pull(movies, Major_Genre) %>%
  unique() %>%
  na.omit()
ratings <- pull(movies, MPAA_Rating) %>%
  unique() %>%
  na.omit()

### functions used in app
scatterplot <- function(df) {
  p <- ggplot(mapping = aes(Rotten_Tomatoes_Rating, IMDB_Rating)) +
    geom_point(data = df %>% filter(selected),  aes(text = Title), size = 2, alpha = 1) +
    geom_point(data = df %>% filter(!selected),  size = .5, alpha = .1)
  ggplotly(p, tooltip = "Title") %>%
    style(hoveron = "fill")
}

### definition of app
ui <- fluidPage(
  titlePanel("IMDB Analysis"),
  selectInput("genres", "Genre", genres),
  checkboxGroupInput("mpaa", "MPAA Rating", ratings, ratings),
  sliderInput("year", "Year", min = min(movies$year), max = max(movies$year), c(1928, 2020), sep = ""),
  plotlyOutput("ratings_scatter")
)

server <- function(input, output) {
 

  output$ratings_scatter <- renderPlotly({
    scatterplot(movies %>%
      mutate(selected = (
        (Major_Genre %in% input$genres) &
        (MPAA_Rating %in% input$mpaa) &
        (year >= input$year[1]) &
        (year <= input$year[2])
      )))
  })
}

app <- shinyApp(ui, server)
```

