---
title: "Untitled"
output: html_document
date: "`r Sys.Date()`"
---

```{r, echo = FALSE}
knitr::opts_chunk$set(warnings = FALSE, message = FALSE)
```

```{r}
library(tidyverse)
library(tidymodels)
theme_set(theme_classic())
```

[Food nutrients] This problem will use PCA to provide a low-dimensional view
of a 14-dimensional nutritional facts [dataset](https://uwmadison.box.com/shared/static/nmgouzobq5367aex45pnbzgkhm7sur63.csv).
The data were originally curated by the USDA.
    
```{r}
nutrients <- read_csv("https://uwmadison.box.com/shared/static/nmgouzobq5367aex45pnbzgkhm7sur63.csv")
View(nutrients)
```
    
a. Define a tidymodels `recipe` that normalizes all nutrient features and run PCA.

```{r}
myRecipe = recipe(~., data = nutrients) %>% 
  update_role(id, name, group, group_lumped, new_role = "id") %>% 
  step_normalize(all_predictors()) %>% 
  step_pca(all_predictors())
  
fit = prep(myRecipe)
```

```{r}
components = tidy(fit, 2)
```

```{r}
scores = juice(fit)
```

b. Visualize the compositions of the top 6 principal components.

```{r}
components %>% 
  filter(component <= "PC6" & str_length(component) == 3) %>% 
  ggplot(aes(value, terms)) + 
  geom_col() +
  facet_wrap(vars(component))
```


c. Compute the average PC2 score for each category of the `group_lumped` column. Order the groups from highest to lowest average PC2 value.

Do these results correspond with the results of the visualization in part (b)?

```{r}
scores %>% 
  group_by(group_lumped) %>% 
  summarize(average = mean(PC2)) %>% 
  arrange(desc(average))
```

d. Visualize the scores of each food item with respect to the first two
principal components. Facet the visualization according to the `group_lumped` column, and order the facets according to part c (highest to lowest PC2 score).

```{r}
ggplot(scores, aes(PC1, PC2)) +
  geom_point() +
  facet_wrap(vars(reorder(group_lumped, -PC2))) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0)
```
