---
title: "Untitled"
output: html_document
date: "`r Sys.Date()`"
---

```{r, echo = FALSE}
knitr::opts_chunk$set(warnings = FALSE, message = FALSE)
```

```{r}
library(tidyverse)
library(shiny)
library(DT) # for shiny's table
theme_set(theme_bw())
```

[Birds Brushed Scatterplot]. The
[`birds`](https://raw.githubusercontent.com/krisrs1128/stat479_s22/main/exercises/data/birds.csv) dataset defined below gives characteristics of the eggs laid by different species of birds. In this problem, you will build an interactive app to compare egg `Asymmetry` and `Ellipticity` across species and genera.

```{r}
birds <- read_csv("https://raw.githubusercontent.com/krisrs1128/stat479_s22/main/exercises/data/birds.csv") |>
    separate(Species, c("genus", "species2"))
```

a. Create a scatterplot of `Asymmetry` vs. `Ellipticity` which supports brushing. Specifically, whenever the user brushes over a set of points in the scatterplot, they should be shown a table of the selected species. Also, change the color of the currently brushed points, so that they stand out more clearly.

```{r}
ggplot(birds, aes(Asymmetry, Ellipticity)) +
  geom_point()
```


```{r}
myScatter = function(myEmphasis) {
    ggplot(birds, aes(Asymmetry, Ellipticity, color = myEmphasis)) +
      geom_point() +
      scale_color_manual(values = c("gray", "black"))
}

ui <- fluidPage(
  plotOutput("scatterplot", brush = "myBrush"),
  DTOutput("ourData")
)

server <- function(input, output) {
  
  emphasisVector = reactiveVal(rep(c(TRUE, FALSE), 1402/2))
  
  observeEvent(
    input$myBrush, {
      brushed_points_dataframe = brushedPoints(birds, input$myBrush, allRows = TRUE)
      about_to_be_emphasis = brushed_points_dataframe$selected_
      emphasisVector(about_to_be_emphasis)
     
    }
    
    )
  
  output$scatterplot = renderPlot(myScatter(emphasisVector()))
  output$ourData = renderDT(birds %>% filter(emphasisVector()))
}

shinyApp(ui, server)
```
