---
title: "Untitled"
output: html_document
date: "`r Sys.Date()`"
---

```{r, echo = FALSE}
knitr::opts_chunk$set(warnings = FALSE, message = FALSE)
```

```{r}
library(tidyverse)
library(tsibble)
library(feasts)
my_theme <- theme_classic() +
  theme(
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14),
    strip.text = element_text(size = 14),
    panel.grid.minor = element_blank()
  )
theme_set(my_theme)
```

[Global Carbon Footprint Network I] This problem studies a dataset of global carbon emissions collected by the Global Carbon Footprint Network. It was the subject of a previous [course project](https://observablehq.com/@wang2274/milestone-4-team-4). We will first quality check and structure the raw data. Then, we'll try answering some time series questions visually.

a. The first few rows of the dataset are printed below. Write down two time-series questions that you think the dataset could help to answer.
    
```{r}
carbon <- read_csv("https://go.wisc.edu/7qx7u1")
head(carbon)
```

1) Is there a global (long-run) trend/pattern in carbon emissions over the entire time-period of this dataset? (Aggregated, individually within each country?)
2) Is there is a cyclical/repeated/seasonal pattern in carbon emissions?

b. Build a `tsibble` object from the data above. Discuss why it can be
useful to convert the original `data.frame` object to this new `tsibble`
class.

```{r}
carbon_tidy = carbon %>% 
  pivot_longer(-c("country", "Country Code"), names_to = "year") %>%
  mutate(year = as.numeric(year)) %>% 
  filter(country != "Switzerland")
  
  
carbon_ts = as_tsibble(carbon_tidy, key = country, index = year)
```


c. Create a line plot and a heatmap visualization of the time series across all countries. What have you learned from your visualization?

```{r}
ggplot(carbon_ts, aes(year, value, group = country)) +
  geom_line() +
  geom_line(data = carbon_ts %>% filter(country == "Bahamas, The"), color = "red", linewidth = 2)
```

```{r}
ggplot(carbon_ts %>% filter(country < "c"), aes(year, country, fill = value)) +
  geom_tile() +
  scale_fill_viridis_c(option = "magma")
```


d. Create a lag plot of the series for the United States. How do you interpret this figure?

```{r}
carbon_ts %>% 
  filter(country == "United States") %>% 
  gg_lag(geom = "point")
```

e. Create a lineplot across all countries after taking a moving average over neighboring years. What are the advantages and disadvantages of smoothing
the original line plot?

```{r}
library(slider)
smoothed_carbon <- carbon_ts |>
  group_by(country) |>
  mutate(smoothed = slide_dbl(value, mean, .before = 5, .after = 5))
```

```{r}
ggplot(smoothed_carbon, aes(year, smoothed, group = country)) +
  geom_line() +
  geom_line(data = smoothed_carbon %>% filter(country == "Bahamas, The"), color = "red", linewidth = 2)
```

