---
title: "Untitled"
output: html_document
date: "`r Sys.Date()`"
---

```{r, echo = FALSE}
knitr::opts_chunk$set(warnings = FALSE, message = FALSE)
```

```{r}
library(tidyverse)
library(caret)
library(DALEX)
library(tidyverse)

th <- theme_minimal() + 
  theme(
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "#f7f7f7"),
    panel.border = element_rect(fill = NA, color = "#0c0c0c", linewidth = 0.6),
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16),
    legend.position = "bottom"
  )
theme_set(th)
```

[Gender Pay Gap] This problem uses CP profiles to investigate the gender gap in a Glassdoor dataset of employee salaries. 

It is helpful to use a model, because it allows us to control for multiple other factors -- a direct plot of salary vs. gender could be criticized as not accounting for confounding variables. The code below trains a gradient boosting machine model on `BasePay` variable (yearly salary in USD), using all potential predictors in the dataset.
    
```{r, warning = FALSE, message = FALSE}
salary <- read_csv("https://github.com/krisrs1128/stat436_s24/raw/main/data/Glassdoor%20Gender%20Pay%20Gap.csv")

x <- salary |>
  select(Gender:Seniority) |>
  mutate(across(where(is.character), as.factor))
y <- salary$BasePay

fit <- train(x, y, method = "gbm", verbose = FALSE)
```

a. Make a basic plot comparing the model's predicted base pay (`predict(fit)`) across Gender. Why does this not show the whole picture?

```{r}
ggplot(salary, aes(Gender, predict(fit))) +
  geom_boxplot()
```

b. Compute aggregate CP profiles grouped by gender and comment on the extent of the gender pay gap. According to the fitted prediction surface, is there more or less of a pay gap at certain ages or levels of seniority?

```{r}
set.seed(2025)

# Compute aggregate CP profiles
explainer = explain(fit, data = x, y = y)
profiles = model_profile(explainer, groups = "Gender")
  
# plot them!
plot(profiles)
```

c. Show the analogous display without aggregating (i.e., `geom = "profiles"`). What is the interpretation of each line in this plot?

```{r}
plot(profiles, geom = "profiles")
```

