---
title: "Untitled"
output: html_document
date: "`r Sys.Date()`"
---

```{r, echo = FALSE}
knitr::opts_chunk$set(warnings = FALSE, message = FALSE)
```

```{r}
library(tidyverse)
library(tidymodels)
theme_set(theme_classic())
```

[Food nutrients] This problem will use PCA to provide a low-dimensional view
of a 14-dimensional nutritional facts [dataset](https://uwmadison.box.com/shared/static/nmgouzobq5367aex45pnbzgkhm7sur63.csv).
The data were originally curated by the USDA.
    
```{r}
nutrients <- read_csv("https://uwmadison.box.com/shared/static/nmgouzobq5367aex45pnbzgkhm7sur63.csv")
View(nutrients)
```
    
a. Define a tidymodels `recipe` that normalizes all nutrient features and run PCA.

```{r}

```


b. Visualize the compositions of the top 6 principal components.

```{r}

```


c. Compute the average PC2 score for each category of the `group_lumped` column. Order the groups from highest to lowest average PC2 value.

Do these results correspond with the results of the visualization in part (b)?

```{r}

```

d. Visualize the scores of each food item with respect to the first two
principal components. Facet the visualization according to the `group_lumped` column, and order the facets according to part c (highest to lowest PC2 score).

```{r}

```
