---
title: "NYC Rentals"
output: html_document
---

```{r}
library(shiny)
library(glue)
library(bslib)
library(tidyverse)

# code to set a clean theme
th <- theme_minimal() + 
  theme(
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "#F7F7F7"),
    panel.border = element_rect(fill = NA, color = "#0c0c0c", size = 0.6),
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16),
    legend.position = "bottom"
  )
theme_set(th)
```
    
[NYC Rentals] In this problem, we'll create a visualization to dynamically
query a [dataset](https://uwmadison.box.com/shared/static/zi72ugnpku714rbqo2og9tv2yib5xped.csv) of Airbnb rentals in Manhattan in 2019. The steps below guide you through the process of building this visualization.

```{r}
rentals <- read_csv("https://uwmadison.box.com/shared/static/zi72ugnpku714rbqo2og9tv2yib5xped.csv") |>
  mutate(price = pmin(price, 1e3))
```

    
a. Make a scatterplot of locations (Longitude vs. Latitude) for all the rentals, colored in by `room_type`.

```{r}
ggplot(rentals, aes(longitude, latitude, color = room_type)) +
  geom_point()
```


b. Design a plot and a dynamic query so that clicking or brushing on the plot updates the points that are highlighted in the scatterplot in (a). For example, you may query a histogram of prices to focus on neighborhoods that are more or less affordable.

```{r}
ggplot(rentals, aes(price)) +
  geom_histogram()
```


```{r}
ui = fluidPage(
  plotOutput("scatterplot", brush = "myBrush"),
  plotOutput("histogram", brush = brushOpts("myBrush", direction = "x"))
)

server = function(input, output) {
  
  emphasisVector = reactiveVal(rep(TRUE, nrow(rentals)))
  
  observeEvent(
    input$myBrush,
    emphasisVector(brushedPoints(rentals, input$myBrush, allRows = TRUE)$selected_)
  )
  
  output$scatterplot = renderPlot(ggplot(rentals, aes(longitude, latitude, color = room_type, alpha = emphasisVector())) +
  geom_point())
  
  output$histogram = renderPlot(
    ggplot(rentals, aes(price)) +
      geom_histogram(fill = "gray") +
      geom_histogram(data = filter(rentals, emphasisVector()), fill = "gray10")
  )
}

shinyApp(ui, server)
```

c. Implement the reverse graphical query. That is, allow the user to update the plot in (b) by brushing over the scatterplot in (a). (Done above)
