---
title: "Untitled"
output: html_document
date: "`r Sys.Date()`"
---

```{r}
knitr::opts_chunk$set(warnings = FALSE, message = FALSE)
```

[Catalan Elections] The `Catalan_elections` dataset contains data on eleven elections in a region of Spain; in each election, the percentage of voters in each of 949 municipalities voting for "Indy" or for "Unionist" are tracked.

```{r}
library(tidyverse)
library(ggridges)
library(ggbeeswarm)
Catalan_elections
```

a. Use a geom from `ggplot2` to visualize: In each of the 11 elections in the dataset, what was the distribution of % voters for Indy in each municipality?

```{r}
Catalan_elections %>% 
  filter(Option == "Indy") %>% 
  ggplot(aes(x = Percent/100, y = factor(Year))) +
  geom_boxplot() +
  # fancy stuff.
  geom_vline(xintercept = 0.5, linetype = "dashed") +
  scale_x_continuous(labels = scales::label_percent()) +
  labs(x = "",
       y = "") +
  theme_classic() +
  theme(panel.grid.major = element_line())
```

b. Use either `geom_density_ridges` from `ggridges` or `geom_beeswarm` from `ggbeeswarm` to answer the same question.

```{r}
Catalan_elections %>% 
  filter(Option == "Indy") %>% 
  ggplot(aes(x = Percent, y = factor(Year))) +
  geom_beeswarm(size = 0.2) 
```

```{r}
Catalan_elections %>% 
  filter(Option == "Indy") %>% 
  ggplot(aes(x = Percent, y = factor(Year))) +
  geom_density_ridges()
```

c. Why did you choose the `geom` you did in part b? How does it improve on the plot from part a?

`geom_density_ridges` uses density as the height for each ridge. This shows the viewer that the *shape* of each ridge can be different; e.g. the 2015 election is much more right skewed than the 1980 election. In a boxplot, the emphasis on that skew is not strong; there is a lot of information that gets lost by just extending a line to the left of the central box.

