---
title: "NYC Rentals"
output: html_document
---

```{r}
library(shiny)
library(glue)
library(bslib)
library(tidyverse)

# code to set a clean theme
th <- theme_minimal() + 
  theme(
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "#F7F7F7"),
    panel.border = element_rect(fill = NA, color = "#0c0c0c", size = 0.6),
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16),
    legend.position = "bottom"
  )
theme_set(th)
```
    
[NYC Rentals] In this problem, we'll create a visualization to dynamically
query a
[dataset](https://uwmadison.box.com/shared/static/zi72ugnpku714rbqo2og9tv2yib5xped.csv)
of Airbnb rentals in Manhattan in 2019. The steps below guide you through the
process of building this visualization.

```{r}
rentals <- read_csv("https://uwmadison.box.com/shared/static/zi72ugnpku714rbqo2og9tv2yib5xped.csv") |>
  mutate(trunc_price = pmin(price, 1e3))
```

    
a. Make a scatterplot of locations (Longitude vs. Latitude) for all the rentals, colored in by `room_type`.

b. Design a plot and a dynamic query so that clicking or brushing on the plot updates the points that are highlighted in the scatterplot in (a). For example, you may query a histogram of prices to focus on neighborhoods that are more or less affordable.

c. Implement the reverse graphical query. That is, allow the user to update the plot in (b) by brushing over the scatterplot in (a).
