---
title: "Untitled"
output: html_document
date: "`r Sys.Date()`"
---

```{r, echo = FALSE}
knitr::opts_chunk$set(warnings = FALSE, message = FALSE)
```

```{r}
library(tidyverse)
th <- theme_minimal() + 
  theme(
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "#f7f7f7"),
    panel.border = element_rect(fill = NA, color = "#0c0c0c"),
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16),
    strip.text = element_text(size = 16),
    legend.position = "bottom"
  )
theme_set(th)
```
    
1. [Polio Clustering] The polio dataset contains weekly counts of new polio cases across each state in the United States, starting as early as 1912. In this question, our goal is to apply $K$-means to group states with similar incidence patterns.

```{r}
polio <- read_csv("https://uwmadison.box.com/shared/static/nm7yku4y9q7ylvz5kbxya3ouj2njd0x6.csv")
head(polio)
```
  
a. Is it necessary to reshape the data before it can be used in the
`kmeans` function? If so, provide the code. If not, provide a brief
explanation.

```{r}

```

b. Let `x` contain the output from part (a). Run $K$-means on `x` with 5 clusters.

```{r}

```

c. Extract and visualize the centroids from the $K$-means fit made in part (b). Explain how to interpret these centroids.

```{r}

```

