---
title: "Untitled"
output: html_document
date: "`r Sys.Date()`"
---

```{r}
library(tidyverse)
library(tidygraph)
library(ggraph)
```

1. [Climate Crisis Youtube recommendations] We will study a
[dataset](https://observablehq.com/@emma-ba/youtube-recommendation-network-on-the-climate-crisis) giving links between Youtube videos related to the climate crisis.  The data were gathered by simulating a user browsing through recommendations, after having initially searched using a set of climate-related seed terms. Each node is a video, which includes features like the number of views and the channel it belongs to. Each edge provides an algorithmically generated recommendation.

```{r}
G = tbl_graph(read_csv("https://github.com/krisrs1128/stat436_s25/raw/main/data/climate-nodes.csv"), read_csv("https://github.com/krisrs1128/stat436_s25/raw/main/data/climate-edges.csv"))
```

a. Visualize the connections between videos as a node-link diagram. Color nodes in by the simulated browser session.

```{r}
ggraph(G, layout = "kk") +
  geom_node_point(aes(color = session)) +
  geom_edge_link()
```

b. Visualize the connections between videos as an adjacency matrix.

```{r}
ggraph(G, layout = "matrix") +
  geom_edge_tile() +
  geom_node_point(aes(color = session), x = 1)
```

c. Compare and contrast the visualizations in parts (a) and (b). What are some of the trade-offs associated with using one view vs. the other?

d. Select a small number of channels of interest (e.g., "The Daily Show with Trevor Noah") For either parts (a) or (b), find a way to visually
distinguish nodes belonging to these channels from the rest.

```{r}
G %>% 
  mutate(Noah = (channel == "The Daily Show with Trevor Noah")) %>% 

ggraph(layout = "kk") +
  geom_edge_link() +
  geom_node_point(aes(color = Noah, size = Noah))
  
```

```{r}
G %>% 
  mutate(Noah = (channel == "The Daily Show with Trevor Noah")) %>%

ggraph(layout = "matrix") +
  geom_edge_tile() +
  geom_node_point(aes(color = Noah, size = Noah), x = 1)
```

