---
title: "Untitled"
output: html_document
date: "`r Sys.Date()`"
---

```{r, echo = FALSE}
knitr::opts_chunk$set(warnings = FALSE, message = FALSE)
```

[Interactive German Traffic] This problem will revisit the previous problem
from an interactive point of view. We will build a visualization that helps
users explore daily traffic patterns across multiple German cities, using
interactivity to help users navigate the collection. We will need additional
features related to the day of the week for each timepoint, created by the
`wday` function below,

```{r}
library(tidyverse)
library(lubridate)
library(shiny)
theme_set(theme_bw())

traffic <- read_csv("https://uwmadison.box.com/shared/static/x0mp3rhhic78vufsxtgrwencchmghbdf.csv") |>
 mutate(day_of_week = wday(date))
```

a. Design and implement a Shiny app that allows users to visualize traffic over time across selected subsets of cities. Make sure that it is possible to view data from more than one city at a time. It is not necessary to label the cities within the associated figure.

```{r}
myCities = c("Holz", "Hasborn", "Kastel")
traffic %>% 
  filter(name %in% myCities) %>% 
ggplot(aes(date, value, group = name)) +
  geom_line()
```

```{r}
traffic_plot = function(dataframe) {
  ggplot(dataframe, aes(date, value, group = name)) +
    geom_line()
}
```


```{r}
ui = fluidPage(
  # way to select cities
  selectInput("user_cities", 
              label = "Select a set of cities",
              choices = unique(traffic$name),
              multiple = TRUE,
              selected = c("Holz", "Hasborn", "Kastel")),
  # the plot itself
  plotOutput("myPlot")
  
)

server = function(input, output) {
  
  chosenRows = reactive({
    traffic %>% 
      filter(name %in% input$user_cities)
  })
  
  output$myPlot = renderPlot({
    traffic_plot(chosenRows())
  })
  
}

shinyApp(ui, server)
```



b. Introduce new inputs to allow users to select a contiguous range of days of the week. For example, the user should have a way of zooming into the samples taken within the Monday - Wednesday range.

```{r}
ui = fluidPage(
  # way to select cities
  selectInput("user_cities", 
              label = "Select a set of cities",
              choices = unique(traffic$name),
              multiple = TRUE,
              selected = c("Holz", "Hasborn", "Kastel")),
  sliderInput("mySlider", 
              label = "Select a time range",
              min = min(traffic$date),
              max = max(traffic$date),
              value = c(min(traffic$date), max(traffic$date))),
  # the plot itself
  plotOutput("myPlot")
  
)

server = function(input, output) {
  
  chosenRows = reactive({
    traffic %>% 
      filter(name %in% input$user_cities,
             date >= input$mySlider[1], date <= input$mySlider[2])
  })
  
  output$myPlot = renderPlot({
    traffic_plot(chosenRows())
  })
  
}

shinyApp(ui, server)
```

