---
title: "Homework 4"
author: Kris Sankaran, adapted by Cameron Jones
date: Summer 2025
  
output: rmdformats::robobook
---

```{r, echo = FALSE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, eval = FALSE)
```

# Instructions

Answer the following pre-written problems in a single R Markdown file. There is no portfolio visualization for this assignment. Knit that file into an .html file and **submit both the R Markdown file and .html file!**

If you fail to submit either the R Markdown file or .html file, your assignment may be subject to penalties. 

A reminder that generative AI and assistance from classmates are allowed in certain capacities on certain problems. That is, on pre-written problems, you may only have *conversations* with classmates about the concepts they test, and not consult their code or any generative AI at all. On portfolio visualizations, you may learn from classmates' code and apply it to your own distinct portfolio visualization, and you may consult generative AI.

# Grading

Each problem will be graded on the following criteria where relevant:

- **Attention to detail in visual design.** Your design choices should go above and beyond defaults, and show critical thinking towards aiding the viewer in understanding the takeaway(s) of the visualization. Strive to make your graphs as polished as possible; especially with choices such as labels, sizes, themes, and colors. 

Instructor feedback on visual design is intended to help the student grow and critically consider visual choices. In particular, plots that show bare minimum effort, or make visual choices that actively hinder the viewer, will be docked points. To receive full credit on visual design, students should "go above and beyond defaults and show critical thinking towards aiding the viewer..." You are encouraged to tinker with subtle visualization elements like themes and layouts, and to explore how to make these changes on your own.

- **Code style.** The code you use to implement your visualizations should be readable and well-organized. Reproducibility is extremely important in the field of data visualization, and messy code is a barrier to it. This also encourages you to organize your code well, which is extremely helpful for the iterative visualization creation process. 

- **Thoughtful explanations, interpretations, and justification.** Writing should be thorough without being overly wordy or jargon-heavy. Writing should draw organically from one's own experiences, opinions, or relevant sources.

- Finally, the **.html file should be well-organized and formatted overall**. Please take care to ensure that in your knitted .html file, we can clearly identify where each problem starts and ends, and things like R warnings and messages are hidden. For example, creating headings for problems titles like "# Heading" is helpful. 

# Pre-Written Problems

## Global Development

**a.** Using the `countries110` dataset from the `rnaturalearth` package, create a map of the world where each country is fully colored (not just the outline) by their value of `INCOME_GRP`. *Hint: `countries110` is compatible with `geom_sf`.*

```{r}
library(rnaturalearth)
data("countries110")
```

**b.** Consider a viewer whose only interest is to find out which countries are in the lowest or highest category of `INCOME_GRP`. Describe a modification you could make to your visualization in part a) to help this viewer identify those countries. Do you think a map is the best type of visualization for this viewer's specific interest? Why or why not?  

## Spotify Time Series

The most streamed song on Spotify in the year 2017 was Ed Sheeran's "Shape of You". The dataset below contains the number of daily streams for this song for each day in various regions. 

```{r}
daily_streams = read_csv("https://uwmadison.box.com/shared/static/hvplyr3jy6vbt7s80lqgfx81ai4hdl0q.csv")
```

**a.** We wish to create a `tsibble` object from the above dataframe. Explain why the following line of code returns an error when you run it. Then, fix the error, converting the dataframe into a tsibble object with multiple time series. *Hint: You only need to add one argument to `tsibble()`.*

```{r}
ts = tsibble(daily_streams, index = "date")
```

**b.** Filter `ts` to just the "global" `region`. Then, make a `gg_season` plot that helps the viewer answer the question: On which day(s) of the week does this song tend to be streamed the most?

## HIV Network

The code below reads in nodes and edges for a graph describing the similarity of 512 HIV samples collected across 32 countries. Two samples are linked if they have similar genetic sequences.

```{r, echo = TRUE}
library(tidygraph)
nodes <- read_csv("https://github.com/krisrs1128/stat436_s25/raw/refs/heads/main/data/hiv_nodes.csv")
edges <- read_csv("https://github.com/krisrs1128/stat436_s25/raw/refs/heads/main/data/hiv_edges.csv")
G <- tbl_graph(nodes, edges)
G
```

*(Grading note: You only need to get the visualizations for this problem functioning, you do NOT need to make improvements to the visual choices.)*

**a.** Create a basic node-link diagram of this dataset.  Then, comment on what the node-link diagram tells you about the overall structure of the network, and what you would expect the adjacency matrix to look like.

**b.** Visualize the adjacency matrix for this dataset. Comment on how prominent features or patterns of the adjacency matrix show up in the node link diagram and vice versa.


## Movie Genres

How do movie genres relate to one another? Is romance + comedy a more common combination than fantasy + adventure? We will use the dataset [here](https://github.com/krisrs1128/stat436_s24/raw/main/data/movie_links.csv) to answer these questions, using a node-link diagram.

*(Grading note: Your improvements to the visual design WILL be assessed for this problem as described in "Grading".)*

**a.** Build a `tbl_graph` from the provided dataset of edges.

**b.** Create a node-link diagram to represent the connections between movies and their genres. Implement a way to visually distinguish between genres and movies.

The genres in this dataset are: "Action", "Adventure", "Sci-Fi", "Comedy", "Crime", "Animation", "Thriller", "Biography", "Drama", "Family", "Horror", "Fantasy", "Romance", "Musical", "Mystery", "Sport", "War", "Western".
