
```{r}
library(tidyverse)
library(tidymodels)
library(rvest)
library(ggrepel)
library(tidytext)
```

# Read in and process NFL data

## Passing Data

Source: https://www.pro-football-reference.com/years/2023/passing.htm

```{r}
passing_raw = html_table(read_html("https://www.pro-football-reference.com/years/2023/passing.htm#passing"))[[1]]
colnames(passing_raw)[27] = "SkYards"

passing_clean = passing_raw %>%
  mutate(GS = as.numeric(GS),
         Player = str_remove(Player, "\\*"),
         Player = str_remove(Player, "\\+")) %>% 
  filter(GS >= 10, Pos == "QB") %>%
  mutate(PassAttPG = as.numeric(Att)/GS,
         PassTDPG = as.numeric(TD)/GS,) %>% 
  select(Player, GS, PassAttPG, "Cmp%", "Y/G", PassTDPG, "Int%", "Sk%") %>% 
  rename(CompletionPerc = "Cmp%",
         PassYardsPG = "Y/G",
         IntPerc = "Int%",
         SackPerc = "Sk%") %>% 
  mutate(PassYardsPA = PassYardsPG/PassAttPG) %>% 
  select(-PassYardsPG)
```

## Rushing Data

Source: https://www.pro-football-reference.com/years/2023/rushing.htm

```{r}
rushing_raw = passing_raw = html_table(read_html("https://www.pro-football-reference.com/years/2023/rushing.htm#rushing"))[[1]]
colnames(rushing_raw) = rushing_raw[1,]

rushing_clean = rushing_raw %>% 
  filter(Pos == "QB") %>%
  mutate(GS = as.numeric(GS),
         Player = str_remove(Player, "\\*"),
         Player = str_remove(Player, "\\+")) %>%
  mutate(RushAttPG = as.numeric(Att)/GS,
         RushYdsPG = as.numeric(Yds)/GS,
         RushYdsPA = RushAttPG/RushYdsPG,
         FumblePG = as.numeric(Fmb)/GS) %>% 
  filter(Player != "Joshua Dobbs" | (Player == "Joshua Dobbs" & Team == "2TM")) %>% 
  select(Player, RushYdsPA, RushAttPG, FumblePG)

```

## Combine Data

```{r}
quarterbacks = left_join(passing_clean, rushing_clean) %>% select(-GS)
```

# Perform PCA with tidymodels

```{r}
pca_rec = recipe(~., data = quarterbacks) %>% 
  update_role(Player, new_role = "id") %>%
  step_normalize(all_predictors()) %>%
  step_pca(all_predictors())

pca_prep = prep(pca_rec)
```

```{r}
components = tidy(pca_prep, 2) %>% 
  mutate(terms = case_match(
    terms,
    "CompletionPerc" ~ "Completion %",
    "FumblePG" ~ "Fumbles Per Game",
    "IntPerc" ~ "Interception %",
    "PassAttPG" ~ "Pass Att. Per Game",
    "PassTDPG" ~ "Pass TD Per Game",
    "PassYardsPA" ~ "Pass Yards Per Attempt",
    "RushAttPG" ~ "Rush Att. Per Game",
    "RushYdsPA" ~ "Rush Yards Per Attempt",
    "SackPerc" ~ "Sack %"
  ))
```

# Determine What The Components Mean

```{r}
components %>% 
  filter(component <= "PC3") %>% 

ggplot(aes(value, reorder_within(terms, abs(value), component))) +
  geom_col() +
  geom_vline(xintercept = 0, linewidth = 1) +
  facet_wrap(~component, nrow = 1, scales = "free_y") +
  scale_y_reordered() +
  theme_classic() +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  ) +
  labs(
    title = "Decomposition of Principal Components",
    subtitle = "NFL Quarterbacks, 2023-24",
    caption = "Minimum 10 Starts. Source: Pro Football Reference",
    x = "",
    y = ""
  )

ggsave("Week7_Think_Tank_Components.png", height = 5, width = 8)
```

Component 1: High Pass TD, completion %, low sack %, high pass yards per attempt, low fumbles per game. Let's call this "Devastating Passer".

Component 2: High rushing attempts but low yards - also low pass attempts. Let's call this "Loves a Short Rush".

Component 3; Lots of passes but high interception rate... "Lots of Passes, Lots of Picks"

# Plot Quarterbacks by Components 

```{r}
scores = juice(pca_prep)

head(scores) 
```

```{r}
ggplot(scores, aes(PC1, PC2, col = PC3)) +
  geom_point() +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  
  geom_text_repel(aes(label = Player), size = 2.5, max.overlaps = 6) +
  
  scale_color_gradient(
    low = "black",
    high = "red", breaks = c(min(scores$PC3), max(scores$PC3)), labels = c("Lowest", "Highest")
  ) +
  
  theme_classic() +
  theme(axis.title = element_text(size = 12),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        legend.title = element_text(size = 8),
        legend.text = element_text(size = 7),
        plot.title = element_text(size = 15),
        plot.subtitle = element_text(size = 10),
        legend.position = "top",
        legend.title.position = "top") +
  

  labs(
    title = "PCA Decomposition of NFL Quarterbacks",
    subtitle = "Minimum 10 Starts. 2023-24",
    caption = "Source: Pro Football Reference",
    x = "Devastating Passer (PC1)",
    y = "Loves a Short Rush (PC2)",
    color = "Lots of Passes, Lots of Picks (PC3)"
  )

ggsave("Week7_Think_Tank_QBs.png", height = 5, width = 6)
```

